---
title: "Class 12: Homework"
author: "Aigerim (PID: 09919142)"
format: pdf
---




>Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes


```{r}
rs <- read.table("rs8067378.txt", row.names = 1)

head(rs)
```

```{r}
summary(rs)
```


# Subset the data for A/A and G/G genotypes. For this `t.test()` can be used, because it is applied to compare the means of two groups and assess whether there is a significant difference between them:

```{r}
expr_A.A <- rs[rs$geno == "A/A", "exp"]
expr_G.G <- rs[rs$geno == "G/G", "exp"]

t_test <- t.test(expr_A.A, expr_G.G)

t_test
```

```{r}
library(ggplot2)

ggplot(rs, aes(x = geno, y = exp, fill = geno)) +
  geom_boxplot() +
  geom_jitter(position = position_jitter(0.2), alpha = 0.5) +
  labs(title = "Expression levels by Genotype", x = "Genotype", y = "Expression") +
  theme_minimal()
```


>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

The boxplot reveals noticeable differences in the median and distribution of expression levels between A/A and G/G genotypes for ORMDL3. This visual observation suggests a potential impact of the SNP on gene expression. The exceptionally small p-value (less than 0.05) from the statistical test provides strong evidence against the null hypothesis, reinforcing the idea that the expression levels are significantly different between the genotypes. Consequently, both the visual examination of the boxplot and the statistical significance of the p-value collectively indicate that the SNP does indeed affect the expression of ORMDL3. 






